Quinn Ziegler - Univariate Exercises 3

```{r getData, eval=TRUE}
link='https://github.com/EvansDataScience/VisualAnalytics_2_tabularData/raw/master/data/eduwa.rda'

#getting the data TABLE from the file in the cloud:
load(file=url(link))
```

```{r dataDimension, eval=TRUE}
#number of rows and columns
dim(eduwa) #nrow(eduwa) ncol(eduwa)

```

This is the list of the 24 columns:

```{r column_Names, eval=TRUE}
names(eduwa)
```


```{r ordinalTable, eval=TRUE}
table(eduwa$High.Grade,exclude = 'nothing')
```


```{r boxplot_gg, eval=TRUE}
library(ggplot2)
# boxplots do not use frequency tables

# as.numeric produces turns levels of the factor into numbers
box1 = ggplot(eduwa, aes(y=as.numeric(High.Grade))) 
box1 = box1 + geom_boxplot() + coord_flip() # to show it horizontally

box1
```


```{r boxplot_axis_labels, eval=TRUE}
# the labels use the original ordinal levels
ordLabels= levels(eduwa$High.Grade)

box2 = box1 + scale_y_continuous(labels=ordLabels,breaks=1:15)
box2
```

Box plots have important statistical information. The beginning and the ending of the box indicates the first (q1) and the third quantile (q75); and the thicker line in the middle represents the median. All those values are clearly visible, but we can retrieve the values like this:

```{r box_quarts, eval=TRUE}
#get positions
# using 'ggplot_build'
pos_q1=     ggplot_build(box2)$data[[1]]$lower
pos_median= ggplot_build(box2)$data[[1]]$middle
pos_q3=     ggplot_build(box2)$data[[1]]$upper

# using
levels(eduwa$High.Grade)[c(pos_q1,pos_median,pos_q3)]
```

From the information retrieved, we know:

* 25% of the public Schools offer at most 5th GRADE. 
* 50% of the public Schools offer at most 8th GRADE.
* 75% of the public Schools offer at most 12th GRADE. Also, 25% of the schools offer at least 12th grade.

We can find these results with a _detailed_ frequency table; that is, instead of using the command _table_ as we did before, we could try a more advanced function:

```{r advancedTable, warning=FALSE,results='asis', eval=TRUE}
library(summarytools)
freq(eduwa$High.Grade,style = 'rmarkdown')
```



<span style="color:red"> Exercise:<br> Make sure our box plot follows the same design approach and include all the elements as in the bar plot for nominal data.
</span>

```{r}
box3 = box2 + labs(title = "50% of schools in Washington state offer at most 8th grade", 
                   y = "Highest Grade Offered",
                   caption = "Source: US Department of Education") +
            theme(panel.background = element_rect(fill = "white", colour = "grey50")) +
            theme(axis.text.y = element_blank()) +
            theme(axis.ticks.y = element_blank())

box4 = box3 + theme(plot.title = element_text(size = 12, face = "bold"),
                    axis.text.x = element_text(size = 8),
                    axis.title.x = element_text(size = 10, vjust = -2),
                    plot.caption = element_text(hjust = 0, vjust = -1)) 

box4
```



### Counting

<span style="color:red"> Exercise:<br> Make a histogram for the variable  FREE LUNCH, and make sure it has all the right elements, and get rid of unnecessary elements.
</span>

```{r, eval=TRUE}
summary(eduwa$Free.Lunch)
```

```{r, eval=TRUE}
eduwaFL <- eduwa[complete.cases(eduwa$Free.Lunch),]
datahistFL <- hist(eduwa$Free.Lunch)
```

```{r, eval=TRUE}
baseFL = ggplot(eduwaFL, aes(x = Free.Lunch))
histFL = baseFL + geom_histogram(binwidth = 50, boundary = 0, fill = "lightblue", color = "grey50")
histFL = histFL + stat_bin(binwidth = 50, aes(label=..count..), 
                  geom = "text",boundary = 0,vjust=-1, size = 2)

histFL
```

```{r, eval=TRUE}
titleFL2 = "Very few schools in Washington state have more than 500 students \nreceiving free lunch"

histFL2 = histFL + labs(title = titleFL2, 
                        x = "Students Receiving Free Lunch",
                        y = "Number of Schools",
                        caption = "Source: US Department of Education")

histFL3 = histFL2 + theme(panel.background = element_rect(fill = "white", colour = "grey50"),
                          plot.title = element_text(size = 12, face = "bold"),
                          axis.title = element_text(size = 10),
                          axis.text = element_text(size = 8),
                          plot.caption = element_text(hjust = 0),
                          axis.title.x = element_text(vjust = -1),
                          axis.title.y = element_text(vjust = 2))

histFL4 = histFL3 + scale_x_continuous(breaks = c(0,250,500,750,1000,1250,1500,1750), limits = c(0,2000)) +
                    scale_y_continuous(breaks = c(0,100,200,300,400,500,600), limits = c(0,600))

histFL4
```




### Measurement

```{r removeMissing2, eval=TRUE}
# strategy 2: 
eduwa4=eduwa[complete.cases(eduwa$Student.Teacher.Ratio),]

tapply(eduwa4$Student.Teacher.Ratio, 
       eduwa4$County, 
       mean)
```


```{r histMeans, eval=TRUE}
# keeping strategy 2: 
meanValues=tapply(eduwa4$Student.Teacher.Ratio, 
                  eduwa4$County, 
                  mean)
```


```{r summaryMeans, eval=TRUE}
summary(meanValues)
```


<span style="color:red"> Exercise:<br> Do some research and make a histogram and a density plot using ggplot for the variable we just used above.
</span>

```{r, eval=TRUE}
stratiodf <- as.data.frame(meanValues)
stratiodf$county <- rownames(stratiodf)
rownames(stratiodf) <- NULL
head(stratiodf)

summary(stratiodf$meanValues)
```


```{r, eval=TRUE}
meanST <- mean(stratiodf$meanValues)
meanST
```


```{r}
baseST = ggplot(stratiodf, aes(x = meanValues))
histST = baseST + geom_histogram(binwidth = 2, boundary = 0, fill = "white", color = "grey50")
histST = histST + stat_bin(binwidth = 2, aes(label=..count..), 
                  geom = "text",boundary = 0,vjust=-0.5, size = 3)

histST
```

```{r, eval=TRUE}
titleST2 = "Washington state counties have an average student-teacher ratio approximately equal \nto 18 students per teacher"

histST2 = histST + labs(title = titleST2, 
                        x = "Average Student-Teacher Ratio",
                        y = "Number of Counties",
                        caption = "Source: US Department of Education")

histST3 = histST2 + theme(panel.background = element_rect(fill = "white", colour = "grey50"),
                          plot.title = element_text(size = 12, face = "bold"),
                          axis.title = element_text(size = 10),
                          axis.text = element_text(size = 8),
                          plot.caption = element_text(hjust = 0),
                          axis.title.x = element_text(vjust = -1),
                          axis.title.y = element_text(vjust = 2))

histST4 = histST3 + scale_x_continuous(breaks = c(10,12,14,16,18,20,22,24), limits = c(10,24)) +
                    scale_y_continuous(breaks = c(0,5,10,15,20), limits = c(0,20))

histST5 = histST4 + geom_vline(aes(xintercept = mean(stratiodf$meanValues), color="mean"), 
                               size = 1, linetype = "dashed")

histST6 = histST5 + theme(legend.title = element_blank()) +
                    scale_color_manual(labels = c("Mean"), values = c("blue"))

histST6

```


```{r, eval=TRUE}

baseSTden <- ggplot(stratiodf, aes(x = meanValues))

denST <- baseSTden + geom_density()

denST

```


```{r, eval=TRUE}
titleden = "Washington state counties have an average student-teacher ratio approximately \nequal to 18 students per teacher"

denST2 = denST + labs(title = titleden, 
                        x = "Average Student-Teacher Ratio",
                        y = "Number of Counties",
                        caption = "Source: US Department of Education")

denST3 = denST2 + theme(panel.background = element_rect(fill = "white", colour = "grey50"),
                          plot.title = element_text(size = 12, face = "bold"),
                          axis.title = element_text(size = 10),
                          axis.text = element_text(size = 8),
                          plot.caption = element_text(hjust = 0, vjust = -2),
                          axis.title.x = element_text(vjust = -1),
                          axis.title.y = element_text(vjust = 2))

denST4 = denST3 + scale_x_continuous(breaks = c(8,10,12,14,16,18,20,22,24,26), limits = c(8,26))

denST5 = denST4 + geom_vline(aes(xintercept = mean(stratiodf$meanValues), color="mean"), 
                               size = 1, linetype = "dashed") +
                  geom_vline(aes(xintercept = median(stratiodf$meanValues), color="median"),
                               size = 1, linetype = "dashed", show.legend = TRUE)

denST6 = denST5 + theme(legend.title = element_blank()) +
                  scale_color_manual(labels = c("Mean", "Median"), values = c("blue", "red"))

denST6
```


```{r, eval=TRUE}
boxST = ggplot(stratiodf, aes(y=meanValues, z="Test")) 

boxST = boxST + geom_boxplot() + coord_flip() + scale_y_continuous(breaks = c(10,15,20,25),
                                                                   limits = c(10,25))

boxST2 = boxST + labs(title = "Two counties have a student-teacher ratio well below the state average", 
                   y = "Average Student-Teacher Ratio",
                   caption = "Source: US Department of Education") +
                   theme(panel.background = element_rect(fill = "white", colour = "grey50")) +
                   theme(axis.text.y = element_blank()) +
                   theme(axis.ticks.y = element_blank()) +
                   theme(axis.title.y = element_blank())

boxST3 = boxST2 + theme(plot.title = element_text(size = 12, face = "bold"),
                    axis.text.x = element_text(size = 8),
                    axis.title.x = element_text(size = 10),
                    plot.caption = element_text(hjust = 0, vjust = -1)) 

boxST4 = boxST3 + annotate("text", x = -0.1, y = 11.5, label = "Outliers are: \nColumbia and Lincoln", size = 3)

boxST4
```

